Finding a closest point in a lattice of Voronoi's first kind

### 2014-01-10 paper.tex
added minor

We are interested in solving the closest lattice point problem for lattices of Voronoi's first kind.  Let $\Lambda \subset \reals^m$ be an $n$ dimensional lattice of Voronoi's first kind with obtuse superbasis $b_1,\dots,b_{n+1}$ and let $y \in \reals^m$.  We want to find $n$ integers $w_1,\dots,w_n$ that minimise
\[
\| y - \sum_{i=1}^n b_i w_i \|^2.
\]
We can equivalently find $n+1$ integers $w_1,\dots,w_{n+1}$ that minimise
\[
\| y - \sum_{i=1}^{n+1} b_i w_i \|^2.
\]
The iterative procedure described in~\eqref{eq:relvectsminimslicer} will be used to do this.  In what follows it is assumed that $y$ lies in the space spanned by the basis vectors $b_1,\dots,b_{n}$.  This assumption is without loss of generality because $x$ is a closest lattice point to $y$ if and only if $x$ is a the closest lattice point to the orthogonalprojection of $y$ into the space spanned by $b_1,\dots,b_{n}$.  Let
\begin{equation}\label{eq:matrxBobtusebasis}
B = (b_1\,\,b_2\,\,\dots\,\,b_{n+1})
\end{equation}
be the matrix with columns given by $b_1,\dots,b_{n+1}$ and let $z \in \reals^{n+1}$ be a column vector such that $y = Bz$.  We now want to find a column vector $w = (w_1,\dots,w_{n+1})^\prime$ of integers such that
\begin{equation}\label{eq:tominimise}
\| B(z  -  w) \|^2
\end{equation}
is minimised.  Define the column vector $u_0 = \floor{z}$ where $\floor{\cdot}$ denotes the largest integer less than or equal to its argument and operates on vectors elementwise. In view of Theorem~\ref{thm:revvecssuperbase} the iterative procedure~\eqref{eq:relvectsminimslicer} to compute a closest lattice point can be written in the form
\begin{align}
x_{k+1} &= B u_{k+1} \label{eq:xseqfirsttype}  \\
u_{k+1} &= u_k + t_k \nonumber \\
t_k &= \arg\min_{t \in \{0,1\}^{n+1}}\| B(z - u_k - t) \|^2, \label{eq:pvecmin}
\end{align}
where $\{0,1\}^{n+1}$ denotes the set of column vectors of length $n+1$ with elements equal to zero or one.  Observe that the procedure is initialised at the lattice point $x_0 = Bu_0 = B\floor{z}$.  This choice of initial lattice point is important.  In Section~\ref{sec:comp-clos-relev} we show how the minimisation over $\{0,1\}^{n+1}$ in~\eqref{eq:pvecmin} can be computed efficiently in $O(n^3)$ operations by computing a minimum cut in an undirected flow network.  In the remainder of this section we prove that this iterative procedure results in a closest lattice point after at most $n$ iterations.  That is, we show that there exists a positive integer $K \leq n$ such that $x_K$ is a closest lattice point to $y = Bz$.

@@

It is necessary to introduce some notation.  For $S$ a subset of the indices $\{1,\dots,n+1\}$ let $\onebf_S$ denote the column vector of length $n+1$ with $i$th element equal to one if $i \in S$ and $0$ otherwise. %Thus, for example, if $p \in \reals^{n+1}$, then $p+\onebf_S$ denotes the vector $p$ with those elements having index from $S$ incremented by one.  
We define the function
\[
\Phi(S, p) = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j)
\]
where $q_{ij} = \dotprod{b_i}{b_j}$ are the Selling parameters.  We denote by $\bar{S}$ the complement of the set of indices $S$, that is $\bar{S} = \{ i =1,\dots,n+1 \mid i \notin S\}$.

@@

Denote by $\min(p)$ and $\max(p)$ the minimum and maximum values obtained by the elements of the vector $p$ and define the function
\[
\rng(p) = \max(p) - \min(p)
\] 
to return the difference between the maximum and minimum of $p$.  %For example, if $p = (2,-1,4) \in \ints^3$ then $\rng(p) = 4 - (-1)=5$.  We might also 
%For example $\rng(2,-1,4)= 4 - (-1) = 5$.  
Observe that $\rng(p)$ cannot be negative and that if $\rng(p) = 0$ then all of the elements of $p$ are equal.  We define the function $\subrng(p)$ to return the largest subset, say $S$, of the indices of $p$ such that $\min\{p_i, i \in S\} - \max\{p_i, i \notin S\} \geq 2.$  If no such subset exists then $\subrng(p)$ is the empty set $\emptyset$.  For example, 
\[
\subrng(2,-1,4) = \{1,3\}, \qquad \subrng(2,1,3) = \emptyset, \quad  \subrng(1,3,1) = \{2\}.
\]  
To make the definition of $\subrng$ clear we give the following alternative and equivalent definition.  Let $p \in \reals^n$ and let $\sigma$ be the permutation of the indices $\{1,\dots,n\}$ that puts the elements of $p$ in ascending order, that is
\[
p_{\sigma(1)} \leq p_{\sigma(2)} \leq \dots \leq p_{\sigma(n)}.
\]  
Let $T$ be the smallest integer from $\{2,\dots,n\}$ such that $p_{\sigma(T)} - p_{\sigma(T-1)} \geq 2$.  If no such integer $T$ exists then $\subrng(p) = \emptyset$.  Otherwise 
\[
\subrng(p) =  \{ \sigma(T), \sigma(T+1), \dots, \sigma(n) \}.
\]
The following straightforward property of $\subrng$ will be useful.

@@

\begin{lemma}\label{lem:decrngpreservesclosestpoints}
If $v \in \ints^{n+1}$ such that $B(\floor{z} + v)$ is a closest lattice point to $y = Bz$, then $B\big(\floor{z} + \decrng(v)\big)$ is also a closest lattice point to $y$.
\end{lemma}
\begin{proof}
The lemma is trivial if $\subrng(v) = \emptyset$ so that $\decrng(v) = v$.  It remains to prove the lemma when $\subrng(v) \neq \emptyset$.  In this case put $S = \subrng(v)$ and put $u = \decrng(v) = v - \onebf_S$.  Let $\zeta = z - \floor{z}$ be the column vector containing the fractional parts of the elements of $z$.  We have $\zeta - u = \zeta - v + \onebf_S$.  Applying part~\ref{eq:lem:decSellingsinc} of Lemma~\ref{lem:decSellings} with $p = \zeta - v$ we obtain
\begin{align}
\|B(\zeta - v)\|^2 - \|B(\zeta - u)\|^2 &= \Phi(S, \zeta-v) \nonumber \\
&= \sum_{i \in S}\sum_{j \notin S}q_{ij}\big(1 + 2(\zeta_i-\zeta_j) - 2(v_i - v_j)\big). \label{eq:sumsumBzBu}
\end{align}
Observe that $\zeta_i =  z_i - \floor{z_i} \in [0,1)$ for all $i=1,\dots,n+1$ and so $-1 < \zeta_i-\zeta_j < 1$ for all $i,j=1,\dots,n+1$.  Also, for $i \in S$ and $j\notin S$ we have 
\[
v_i - v_j \geq \min\{ v_i, i \in S\} - \max\{v_j, j \notin S\} \geq 2
\]
by definition of $\subrng(v) = S$.  Thus,
\[
1 + 2(\zeta_i-\zeta_j) - 2(v_i - v_j) < 1 + 2 - 4 = -1 < 0 \qquad \text{for $i \in S$ and $j \notin S$}.
\]
Substituting this inequality into~\eqref{eq:sumsumBzBu} and using that $q_{ij} \leq 0$ (the obtuse condition~\eqref{eq:obtusecond}) we find that
\[
\|B(z - \floor{z} - v)\|^2 - \|B(z - \floor{z} - u)\|^2 \geq 0
\]
and so $B(\floor{z} + u) = B\big(\floor{z} + \decrng(v)\big)$ is a closest lattice point to $y = Bz$ whenever $B(\floor{z} + v)$ is.
\end{proof}

@@

\begin{lemma}\label{eq:integergreaterless}
Let $h \in \{0,1\}^{n+1}$ and $v \in \ints^{n+1}$.  Suppose that
$h_i = 0$ whenever $v_i = \min(v)$ and that $h_i = 1$ whenever $v_i = \max(v)$.  Then
\[
h_i - h_j \geq v_i - v_j
\]
when either $v_i = \min(v)$ or $v_j = \max(v)$.
\end{lemma}
\begin{proof}
If $v_i = \min(v)$ then $h_i = 0$ and we need only show that $-h_j \geq \min(v) - v_j$ for all $j$.  If $v_j = \min(v)$ then $h_j = 0 = \min(v) - v_j$ and the results holds.  Otherwise if $v_j > \min(v)$ then $-h_j \geq -1 \geq \min(v) - v_j$ because $\min(v)$ and $v_j$ are integers.

Now, if $v_j = \max(v)$ then $h_j = 1$ and we need only show that $h_i - 1 \geq v_i - \max(v)$ for all $i$.  If $v_i = \max(v)$ then $h_i = 1$ and $h_i - 1 = 0 = v_i - \max(v)$.  Otherwise if $v_i < \max(v)$ then $h_i - 1 \geq -1 \geq v_i - \max(v)$.
\end{proof}

@@

As stated already $h_i = 0$ whenever $v_i = \min(v)$ and $h_i = 1$ whenever $v_i = \max(v)$.  It follows by Lemma~\ref{eq:integergreaterless} that
\begin{equation}\label{eq:hijvijieq}
h_{i} - h_j \geq v_{i} - v_j
\end{equation}
when either $v_i = \min(v)$ or $v_j = \max(v)$.  Since $v_i = \min(v)$ for $i \in T$ and $v_j = \max(v)$ for $j \in S$ the inequality~\eqref{eq:hijvijieq} holds whenever $i \in T$ or $j \in S$.

Put $r = z - u - h$.  By~\eqref{eq:hijvijieq} we have $r_i - r_j \leq p_{i} - p_{j}$ when either $i \in T$ or $j \in S$.  Now, since $q_{ij} \leq 0$ for $i \neq j$,
\[
\Phi(S,r) = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2r_i - 2r_j) \geq \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j) = \Phi(S,p).
\]
Similarly
\[
\Phi(\bar{T},r) = \sum_{i \notin T}\sum_{j \in T}q_{ij}(1 + 2r_i - 2r_j) \geq \sum_{i \notin T}\sum_{j \in T}q_{ij}(1 + 2p_i - 2p_j) = \Phi(\bar{T},p).
\]\
Now
\begin{align*}
\|B(z-u-h)\|^2 - \|B(z-u-g)\|^2 &= \|Br\|^2 - \|B(r+\onebf_S - \onebf_T)\|^2 \\
&= \Phi(S,r) + \Phi(\bar{T},r) + \sum_{i\in S}\sum_{j \in S} q_{ij} \\
&\leq \Phi(S,p) + \Phi(\bar{T}, p) + \sum_{i\in S}\sum_{j \in S} q_{ij} < 0. 
\end{align*}
as a result of~\eqref{eq:BpBpdecincineq}.  However, $d \in \{0,1\}^{n+1}$ and so
\[
\|B(z-u-g)\|^2 > \|B(z-u-h)\|^2 \geq \min_{t \in \{0,1\}^{n+1}}\|B(z - u - t)\|^2
\]
contradicting~\eqref{eq:qismin}.  Thus, $B(u+g+w)$ is a closest lattice point to $y$ and, since $\rng(w) = \ell-1$, the lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
 \end{proof}

>>>>

We are interested in solving the closest lattice point problem for lattices of Voronoi's first kind.  Let $\Lambda \subset \reals^m$ be an $n$ dimensional lattice of Voronoi's first kind with obtuse superbasis $b_1,\dots,b_{n+1}$ and let $y \in \reals^m$.  We want to find $n$ integers $w_1,\dots,w_n$ that minimise
\[
\| y - \sum_{i=1}^n b_i w_i \|^2.
\]
We can equivalently find $n+1$ integers $w_1,\dots,w_{n+1}$ that minimise
\[
\| y - \sum_{i=1}^{n+1} b_i w_i \|^2.
\]
The iterative procedure described in~\eqref{eq:relvectsminimslicer} will be used to do this.  In what follows it is assumed that $y$ lies in the space spanned by the basis vectors $b_1,\dots,b_{n}$.  This assumption is without loss of generality because $x$ is a closest lattice point to $y$ if and only if $x$ is a the closest lattice point to the orthogonal projection of $y$ into the space spanned by $b_1,\dots,b_{n}$.  Let
\begin{equation}\label{eq:matrxBobtusebasis}
B = (b_1\,\,b_2\,\,\dots\,\,b_{n+1})
\end{equation}
be the matrix with columns given by $b_1,\dots,b_{n+1}$ and let $z \in \reals^{n+1}$ be a column vector such that $y = Bz$.  We now want to find a column vector $w = (w_1,\dots,w_{n+1})^\prime$ of integers such that
\begin{equation}\label{eq:tominimise}
\| B(z  -  w) \|^2
\end{equation}
is minimised.  Define the column vector $u_0 = \floor{z}$ where $\floor{\cdot}$ denotes the largest integer less than or equal to its argument and operates on vectors elementwise. In view of Theorem~\ref{thm:revvecssuperbase} the iterative procedure~\eqref{eq:relvectsminimslicer} to compute a closest lattice point can be written in the form
\begin{align}
x_{k+1} &= B u_{k+1} \label{eq:xseqfirsttype}  \\
u_{k+1} &= u_k + t_k \nonumber \\
t_k &= \arg\min_{t \in \{0,1\}^{n+1}}\| B(z - u_k - t) \|^2, \label{eq:pvecmin}
\end{align}
where $\{0,1\}^{n+1}$ denotes the set of column vectors of length $n+1$ with elements equal to zero or one.  The procedure is initialised at the lattice point $x_0 = Bu_0 = B\floor{z}$.  This choice of initial lattice point is important.  In Section~\ref{sec:comp-clos-relev} we show how the minimisation over $\{0,1\}^{n+1}$ in~\eqref{eq:pvecmin} can be computed efficiently in $O(n^3)$ operations by computing a minimum cut in an undirected flow network.  In the remainder of this section we prove that this iterative procedure results in a closest lattice point after at most $n$ iterations.  That is, we show that there exists a positive integer $K \leq n$ such that $x_K$ is a closest lattice point to $y = Bz$.

@@

It is necessary to introduce some notation.  For $S$ a subset of indices $\{1,\dots,n+1\}$ let $\onebf_S$ denote the column vector of length $n+1$ with $i$th element equal to one if $i \in S$ and $0$ otherwise. %Thus, for example, if $p \in \reals^{n+1}$, then $p+\onebf_S$ denotes the vector $p$ with those elements having index from $S$ incremented by one.  
For $S \subseteq \{1,\dots,n+1\}$ and $p \in \reals^{n+1}$ we define the function
\[
\Phi(S, p) = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j)
\]
where $q_{ij} = \dotprod{b_i}{b_j}$ are the Selling parameters from~\eqref{eq:obtusecond}.  We denote by $\bar{S}$ the complement of the set of indices $S$, that is $\bar{S} = \{ i \in \{1,\dots,n+1\} \mid i \notin S\}$.

@@

Denote by $\min(p)$ and $\max(p)$ the minimum and maximum values obtained by the elements of the vector $p$ and define the function
\[
\rng(p) = \max(p) - \min(p).
\] 
%to return the difference between the maximum and minimum of $p$.  %For example, if $p = (2,-1,4) \in \ints^3$ then $\rng(p) = 4 - (-1)=5$.  We might also 
%For example $\rng(2,-1,4)= 4 - (-1) = 5$.  
Observe that $\rng(p)$ cannot be negative and that if $\rng(p) = 0$ then all of the elements of $p$ are equal.  We define the function $\subrng(p)$ to return the largest subset, say $S$, of the indices of $p$ such that $\min\{p_i, i \in S\} - \max\{p_i, i \notin S\} \geq 2.$  If no such subset exists then $\subrng(p)$ is the empty set $\emptyset$.  For example, 
\[
\subrng(2,-1,4) = \{1,3\}, \qquad \subrng(2,1,3) = \emptyset, \quad  \subrng(1,3,1) = \{2\}.
\]  
To make the definition of $\subrng$ clear we give the following alternative and equivalent definition.  Let $p \in \reals^n$ and let $\sigma$ be the permutation of the indices $\{1,\dots,n\}$ that puts the elements of $p$ in ascending order, that is
\[
p_{\sigma(1)} \leq p_{\sigma(2)} \leq \dots \leq p_{\sigma(n)}.
\]  
Let $T$ be the smallest integer from $\{2,\dots,n\}$ such that $p_{\sigma(T)} - p_{\sigma(T-1)} \geq 2$.  If no such integer $T$ exists then $\subrng(p) = \emptyset$.  Otherwise 
\[
\subrng(p) =  \{ \sigma(T), \sigma(T+1), \dots, \sigma(n) \}.
\]
The following straightforward property of $\subrng$ will be useful.

@@

\begin{lemma}\label{lem:decrngpreservesclosestpoints}
If $v \in \ints^{n+1}$ such that $B(\floor{z} + v)$ is a closest lattice point to $y = Bz$, then $B\big(\floor{z} + \decrng(v)\big)$ is also a closest lattice point to $y$.
\end{lemma}
\begin{proof}
The lemma is trivial if $\subrng(v) = \emptyset$ so that $\decrng(v) = v$.  It remains to prove the lemma when $\subrng(v) \neq \emptyset$.  In this case put $S = \subrng(v)$ and put 
\[
u = \decrng(v) = v - \onebf_S.
\] 
Let $\zeta = z - \floor{z}$ be the column vector containing the fractional parts of the elements of $z$.  We have $\zeta - u = \zeta - v + \onebf_S$.  Applying part~\ref{eq:lem:decSellingsinc} of Lemma~\ref{lem:decSellings} with $p = \zeta - v$ we obtain
\begin{align}
\|B(\zeta - v)\|^2 - \|B(\zeta - u)\|^2 &= \Phi(S, \zeta-v) \nonumber \\
&= \sum_{i \in S}\sum_{j \notin S}q_{ij}\big(1 + 2(\zeta_i-\zeta_j) - 2(v_i - v_j)\big). \label{eq:sumsumBzBu}
\end{align}
Observe that $\zeta_i =  z_i - \floor{z_i} \in [0,1)$ for all $i=1,\dots,n+1$ and so $-1 < \zeta_i-\zeta_j < 1$ for all $i,j=1,\dots,n+1$.  Also, for $i \in S$ and $j\notin S$ we have 
\[
v_i - v_j \geq \min\{ v_i, i \in S\} - \max\{v_j, j \notin S\} \geq 2
\]
by definition of $\subrng(v) = S$.  Thus,
\[
1 + 2(\zeta_i-\zeta_j) - 2(v_i - v_j) < 1 + 2 - 4 = -1 < 0 \qquad \text{for $i \in S$ and $j \notin S$}.
\]
Substituting this inequality into~\eqref{eq:sumsumBzBu} and using that $q_{ij} \leq 0$ for $i \neq j$ (the obtuse condition~\eqref{eq:obtusecond}) we find that
\[
\|B(z - \floor{z} - v)\|^2 - \|B(z - \floor{z} - u)\|^2 \geq 0.
\]
It follows that $B(\floor{z} + u) = B\big(\floor{z} + \decrng(v)\big)$ is a closest lattice point to $y = Bz$ whenever $B(\floor{z} + v)$ is.
\end{proof}

@@

\begin{lemma}\label{eq:integergreaterless}
Let $h \in \{0,1\}^{n+1}$ and $v \in \ints^{n+1}$.  Suppose that
$h_i = 0$ whenever $v_i = \min(v)$ and that $h_i = 1$ whenever $v_i = \max(v)$.  Then
\[
h_i - h_j \leq v_i - v_j
\]
when either $v_i = \max(v)$ or $v_j = \min(v)$.
\end{lemma}
\begin{proof}
If $v_i = \max(v)$ then $h_i = 1$ and we need only show that $1-h_j \leq \max(v) - v_j$ for all $j$.  If $v_j = \max(v)$ then $h_j = 0 = \min(v) - v_j$ and the results holds.  Otherwise if $v_j < \max(v)$ then $1-h_j \leq 1 \leq \max(v) - v_j$ because $\max(v)$ and $v_j$ are integers.

Now, if $v_j = \min(v)$ then $h_j = 0$ and we need only show that $h_i \leq v_i - \min(v)$ for all $i$.  If $v_i = \min(v)$ then $h_i = 0 = v_i - \min(v)$ and the results holds.  Otherwise if $v_i > \min(v)$ then $h_i \leq 1 \leq v_i - \min(v)$ because $\min(v)$ and $v_i$ are integers.
\end{proof}

@@

As stated already $h_i = 0$ whenever $v_i = \min(v)$ and $h_i = 1$ whenever $v_i = \max(v)$.  It follows from Lemma~\ref{eq:integergreaterless} that
\begin{equation}\label{eq:hijvijieq}
h_{i} - h_j \leq v_{i} - v_j
\end{equation}
when either $v_i = \max(v)$ or $v_j = \min(v)$.  Since $v_i = \max(v)$ for $i \in S$ and $v_j = \min(v)$ for $j \in T$ the inequality~\eqref{eq:hijvijieq} holds whenever $i \in S$ or $j \in T$.

Put $r = z - u - h$.  By~\eqref{eq:hijvijieq} we have $r_i - r_j \geq p_{i} - p_{j}$ when either $i \in S$ or $j \in T$.  Now, since $q_{ij} \leq 0$ for $i \neq j$,
\[
\Phi(S,r) = \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2r_i - 2r_j) \leq \sum_{i \in S}\sum_{j \notin S}q_{ij}(1 + 2p_i - 2p_j) = \Phi(S,p).
\]
Similarly,
\[
\Phi(\bar{T},r) = \sum_{i \notin T}\sum_{j \in T}q_{ij}(1 + 2r_i - 2r_j) \leq \sum_{i \notin T}\sum_{j \in T}q_{ij}(1 + 2p_i - 2p_j) = \Phi(\bar{T},p).
\]\
Now
\begin{align*}
\|B(z-u-h)\|^2 - \|B(z-u-g)\|^2 &= \|Br\|^2 - \|B(r+\onebf_S - \onebf_T)\|^2 \\
&= \Phi(S,r) + \Phi(\bar{T},r) + \sum_{i\in S}\sum_{j \in S} q_{ij} \\
&\leq \Phi(S,p) + \Phi(\bar{T}, p) + \sum_{i\in S}\sum_{j \in S} q_{ij} < 0 
\end{align*}
as a result of~\eqref{eq:BpBpdecincineq}.  However, $d \in \{0,1\}^{n+1}$ and so
\[
\|B(z-u-g)\|^2 > \|B(z-u-h)\|^2 \geq \min_{t \in \{0,1\}^{n+1}}\|B(z - u - t)\|^2
\]
contradicting~\eqref{eq:qismin}.  Thus, $B(u+g+w)$ is a closest lattice point to $y$ and, since $\rng(w) = \ell-1$, the lattice point $B(u+g)$ is $(\ell-1)$-close to $y$.
 \end{proof}
